Perturbing SU($n$)-symmetric interactions

### 2020-03-05 collective_sun/sun_perturbation.tex
removed obsolete sections and fixed references appropriately

We consider an array of $N$ multilevel spins with SU($n$)-symmetric
interactions that can be written in the form
\begin{align}
  H_0 = \sum_{p<q} h_{pq} \Pi_{pq},
  &&
  \Pi_{pq} \equiv \sum_{\mu,\nu} s_{\mu\nu}^{(p)} s_{\nu\mu}^{(q)},
  \label{eq:ints}
\end{align}
where $p,q$ index individual spins; $\mu,\nu$ index states in an
orthonormal basis for the $n$-dimensional Hilbert space $\H_n$ of a
single spin; $h_{pq}$ are scalar coefficients; the operator
$s_{\mu\nu}^{(p)}\equiv\op{\mu}{\nu}_p$ flips the state of spin $p$ to
$\ket\mu$ from $\ket\nu$; and the operator $\Pi_{pq}$ permutes the
states of spins $p$ and $q$.  The permutation operator $\Pi_{pq}$ is a
multilevel generalization of the SU(2)-symmetric spin interaction
$\v s_p\c\v s_q$ for $\v s=\p{s_\x,s_\y,s_\z}$, with
$s_\alpha=\sigma_\alpha/2$ and $\sigma_\alpha$ a Pauli matrix for an
individual spin.

@@

If the coefficients $h_{pq}$ of the interaction Hamiltonian $H_0$ are
all negative, then the ground-state manifold $\M_0$ of the interaction
Hamiltonian $H_0$ consists of permutationally symmetric states that
are simultaneous $+1$ eigenstates of all permutation operators
$\Pi_{pq}$.  We will adopt the restriction that all $h_{pq}<0$
throughout these notes, but note that this restriction can be relaxed
to the assumption that the initial state of all spins is
permutationally symmetric (e.g.~a spin-polarized state).  We will also
assume throughout these notes that the permutationally symmetric
manifold $\M_0$ is gapped away from all other states by an interaction
energy difference that is large compared to any coupling between
$\M_0$ and its orthogonal complement.  Power-law couplings of the form
$h_{pq}=-h/\abs{p-q}^\alpha$ on a $D$-dimensional lattice, for
example, always yield a non-vanishing spectral gap when $\alpha\le D$
(see Appendix \ref{sec:gap}).

@@

To compute the effective Hamiltonian induced by the single-body
perturbation $\V_1$, we use the permutational symmetry of the
ground-state manifold to simplify
\begin{align}
  H_1^{(1)} = \sum_{X,p} v_{Xp} \P_0 X_p \P_0
  \EQPS \sum_{X,p} v_{Xp} X_1
  \EQPS \sum_X \bar v_X \col{X},
  \label{eq:H_1_1}
\end{align}
where $\EQPS$ denotes equality under a restriction to the
permutationally symmetric manifold $\M_0$, and
\begin{align}
  \bar v_X \equiv \f1N \sum_p v_{Xp},
  &&
  \col{X} \equiv \sum_p X_p,
\end{align}
are respectively the mean value in $\v v_X$ and the collective version
of $X$.  In order to compute the second order effective Hamiltonian
$H_1^{(2)}$, we first choose an arbitrary permutationally symmetric
state $\ket\psi\in\M_0$ and expand (see Appendix \ref{sec:H_V1_psi}):
\begin{align}
  H_0 \V_1 \ket\psi
  = E_0 \V_1 \ket\psi
  + \sum_X \sum_p \p{\sum_q h_{pq} v_{Xq} - h_p v_{Xp}} X_p
  \ket\psi,
  &&
  h_p \equiv \sum_q h_{pq},
  \label{eq:H_V1_psi}
\end{align}
where we define $h_{pq}$ for all $p,q$ (as opposed to only $p<q$) by
enforcing $h_{pq}=h_{qp}$ and $h_{pp}=0$.  We thus find that a
perturbation $\V_1^\Delta\ket\psi$ (indexed by $\Delta$) is an
eigenvector of the interaction Hamiltonian $H_0$ if its coefficient
vectors $\v v_X^\Delta$ all satisfy the eigenvalue equation
\begin{align}
  \p{\m h - \diag\v h}\c\v v = \Delta \v v,
  \label{eq:cond_1}
\end{align}
where $\m h\equiv\sum_{p,q}h_{pq}\op{p}{q}$ is a matrix of all
couplings $h_{pq}$; the vector $\v h\equiv\sum_{p,q}h_{pq}\ket{p}$ is
the sum of all columns of $\m h$; and $\diag\v h$ is a matrix with
$\v h$ on the diagonal and zeroes everywhere else.  Solving the
general eigenvalue problem in \eqref{eq:cond_1} requires diagonalizing
the $N\times N$ matrix $\m h-\diag\v h$.  When the interaction
Hamiltonian $H_0$ is translationally invariant, however, this
eigenvalue problem can be solved analytically; we provide the
corresponding solution in Appendix \ref{sec:trans_inv}.

If we {\it construct} a perturbation $\V_1^\Delta$ with coefficients
$\v v_X^\Delta$ that satisfy \eqref{eq:cond_1} with eigenvalue
$\Delta$, then
\begin{align}
  H_0 \V_1^\Delta \ket\psi = \p{E_0 + \Delta} \V_1^\Delta \ket\psi.
\end{align}
Interestingly, the energy of the state $\V_1^\Delta\ket\psi$ depends
only on the coefficients $\v v_X^\Delta$, and is entirely independent
of the state $\ket\psi$ or the choice of a trace-zero single-spin
operator $X$ used to build $\V_1^\Delta$.  Finding operators
$\V_1^\Delta$ that generate eigenvectors of the interaction
Hamiltonian $H_0$ when they act on permutationally symmetric states
$\ket\psi\in\M_0$ thus reduces to finding eigenvectors of the
interaction matrix $\m h-\diag\v h$.

% If the permutationally symmetric manifold $\M_0$ is gapped away from
% all other states, then then a vector $\V_1^\Delta\ket\psi$ with
% $\Delta=0$ must lie within the permutationally symmetric manifold
% $\M_0$, which implies that the operator $\V_1^\Delta$ preserves the
% permutational symmetry of $\M_0$.  Indeed, a constant vector
% $\v\1\equiv\p{1,1,1,\cdots}$ of ones satisfies the condition in
% \eqref{eq:cond_1} with eigenvalue $0$, which implies that all
% vectors $\v v$ satisfying \eqref{eq:cond_1} with $\Delta\ne0$ must
% be orthogonal to $\v\1$, and therefore mean-zero.

% We now return to the task of computing the second-order effective
% Hamiltonian $H_1^{(2)}$.
Any coefficient vector $\v v_X$ can be expanded into its projections
$\v v_X^\Delta$ onto the eigenspace of vectors satisfying
\eqref{eq:cond_1} with eigenvalue $\Delta$,
i.e.~$\v v_X = \sum_\Delta \v v_X^\Delta$, which also allows us to
expand
\begin{align}
  \V_1 = \sum_\Delta \V_1^\Delta,
  &&
  \V_1^\Delta \equiv \sum_X \v v_X^\Delta \c \v X,
\end{align}
where each operator $\V_1^\Delta$ generates a state with definite
interaction energy $\Delta$ above that of the permutationally
symmetric manifold $\M_0$.  We can therefore simplify
\begin{align}
  H_1^{(2)}
  = - \P_0 \V_1 \E \V_1 \P_0
  = - \sum_{\Delta\ne0} \f1{\Delta}
  \P_0 \V_1^\Delta \P_\Delta \V_1^\Delta \P_0
  = - \sum_{\Delta\ne0} \f1{\Delta} \P_0 \p{\V_1^\Delta}^2 \P_0,
\end{align}
where the product $\P_0 \p{\V_1^\Delta}^2 \P_0$ is worked out in
Appendix \ref{sec:PXYP}.  In total, we find that
\begin{align}
  H_1^{(2)}
  \EQPS \f1{N\p{N-1}} \sum_{X,Y} \sum_{\Delta\ne0}
  \f{\v v_X^\Delta\c\v v_Y^\Delta}{\Delta}
  \p{\col{X}\,\col{Y} - N \col{XY}},
  &&
  \col{Z} \equiv \sum_p Z_p.
  \label{eq:H_1_2}
\end{align}
where $\EQPS$ denotes equality under a restriction to the
permutationally symmetric manifold $\M_0$; and $\v v_X^\Delta$ is the
projection of $\v v_X$ onto the $\Delta$-eigenspace of the matrix
$\m h-\diag\v h$.

@@

The first-order effective Hamiltonian $H_2^{(1)}$ induced on the
permutationally symmetric manifold $\M_0$ by the two-body perturbation
$\V_2$ in \eqref{eq:perturbations} is simply the restriction of $\V_2$
onto $\M_0$:
\begin{align}
  H_2^{(1)} = \sum_O \sum_{p<q} w_{Opq} \P_0 O_{pq} \P_0
  \EQPS \sum_O \sum_{p<q} w_{Opq} O_{1,2}
  \EQPS \sum_O \bar w_O \col{O},
\end{align}
where
\begin{align}
  \bar w_O \equiv {N \choose 2}^{-1} \sum_{p<q} w_{Opq},
  &&
  \col{O} \equiv \sum_{p<q} O_{pq},
\end{align}
are an average of $w_{Opq}$ and a collective version of the two-spin
operator $O_{pq}$.  If the two-spin operator
$O=X\otimes Y+Y\otimes X$, then
\begin{align}
  H_2^{(1)} \EQPS \f12 \sum_{\p{X,Y}} \bar w_{\p{X,Y}}
  \p{\col{X}\,\col{Y} + \col{Y}\,\col{X} - \col{XY} - \col{YX}}.
\end{align}
In order to compute the second-order effective Hamiltonian
$H_2^{(2)}$, as before we pick an arbitrary permutationally symmetric
state $\ket\psi\in\M_0$ and expand (see Appendix \ref{sec:H_VM_psi})
\begin{align}
  H_0 \V_2 \ket\psi
  = E_0 \V_2 \ket\psi
  + \sum_O \sum_{p<q}
  \sp{\sum_k \p{h_{pk} w_{Okq} + h_{qk} w_{Opk}}
  - \bar h_{pq} w_{Opq}}
  O_{pq} \ket\psi,
  \label{eq:H_V2_psi}
\end{align}
where to keep notation compact, we define
\begin{align}
  \bar h_{pq} \equiv h_p + h_q - 2 h_{pq},
  &&
  h_p \equiv \sum_q h_{pq},
\end{align}
and we define $w_{Opq}$ for all $p,q$ (as opposed to only $p<q$) by
enforcing $w_{Opq}=w_{Oqp}$ and $w_{Opp}=0$.  We thus find that the
vector $\V_2^\Delta\ket\psi$ (indexed by $\Delta$) is an eigenvector
of the interaction Hamiltonian $H_0$ if its coefficients
$\v{\m w}_O^\Delta$ all satisfy the eigenvalue equation
\begin{align}
  \check{\m h} \c \v{\m w} = \Delta \v{\m w},
  \label{eq:cond_2}
\end{align}
where
\begin{align}
  \check{\m h}
  \equiv \sum_{p<q} \ket{\set{p,q}} \sp{\sum_{k\notin\set{p,q}}
    \p{h_{pk} \bra{\set{k,q}} + h_{qk} \bra{\set{p,k}}}
    - \bar h_{pq} \bra{\set{p,q}}}.
  \label{eq:h_super_mat}
\end{align}
We explicitly exclude $k\in\set{p,q}$ in \eqref{eq:h_super_mat}
because these terms vanish in \eqref{eq:H_V2_psi} from the fact that
$h_{kk}=w_{Okk}=0$.  Solving the general eigenvalue problem in
\eqref{eq:cond_2} requires diagonalizing the
${N \choose 2}\times{N \choose 2}\sim N^2\times N^2$ matrix
$\check{\m h}$.  When both he interaction Hamiltonian $H_0$
translationally invariant, however, this eigenvalue problem can be
reduced to that of diagonalizing an $\sim N\times N$ matrix; we
discuss this reduction in Appendix \ref{sec:trans_inv}.

If we {\it construct} a perturbation $\V_2^\Delta$ with coefficients
$\v{\m w}_O^\Delta$ that satisfy \eqref{eq:cond_2} with eigenvalue
$\Delta$, then
\begin{align}
  H_0 \V_2^\Delta \ket\psi = \p{E_0 + \Delta} \V_2^\Delta \ket\psi,
\end{align}
which implies that $\V_2^\Delta\ket\psi$ is a state of definite energy
that depends only on the coefficients $\m w_O^\Delta$.  Similarly to
the case of single-body perturbations, we can decompose any vector
$\v{\m w}_O$ into its projections $\v{\m w}_O^\Delta$ onto the
eigenspace of vectors satisfying \eqref{eq:cond_2} with eigenvalue
$\Delta$, i.e.~$\v{\m w}_O=\sum_\Delta\v{\m w}_O^\Delta$, which allows
us to expand
\begin{align}
  \V_2 = \sum_\Delta \V_2^\Delta,
  &&
  \V_2^\Delta \equiv \sum_O \v{\m w}_O^\Delta \c \v{\m O},
\end{align}
and in turn write
\begin{align}
  H_2^{(2)} = - \P_0 \V_2 \E \V_2 \P_0
  = -\sum_{\Delta\ne0} \f1\Delta \P_0 \p{\V_2^\Delta}^2 \P_0,
\end{align}
where the product $\P_0 \p{\V_2^\Delta}^2 \P_0$ is simplified in
Appendix \ref{sec:POQP}.  In the special case that
\begin{align}
  \V_2 = \sum_{p<q} w_{pq} s_\z^{(p)} s_\z^{(q)},
\end{align}
where $s_\z\equiv\p{\op\up-\op\dn}/2$ is an SU(2) spin-$z$ operator,
the second-order effective Hamiltonian $H_2^{(2)}$ takes the form
\begin{align}
  H_2^{(2)} \simeq_\PS W_4 S_\z^4 - W_2 S_\z^2,
\end{align}
where $\simeq_\PS$ denotes equality up to scalar terms and a
restriction to the permutationally symmetric manifold;
$S_\z\equiv\sum_p s_\z^{(p)}$ is a collective SU(2) spin operator; and
$W_k$ are scalar coefficients.  In order to expand the coefficients
$W_k$, we define
\begin{align}
  \v{\m w}=\sum_{p<q} w_{pq} \ket{\set{p,q}},
  &&
  \v w \equiv \sum_{p,q} w_{pq} \ket{p},
\end{align}
where $w_{pq}$ is defined for all $p,q$ (as opposed to only $p<q$) by
enforcing $w_{pq}=w_{qp}$ and $w_{pp}=0$.  The vector $\v w$ is the
sum of all columns of the matrix
$\m w\equiv\sum_{p,q}w_{pq}\op{p}{q}$.  We then denote the projection
of $\v{\m w}$ onto the $\Delta$-eigenspace of the matrix
$\check{\m h}$ defined in \eqref{eq:h_super_mat} by $\v{\m w}_\Delta$,
and define
\begin{align}
  w_{\t{columns}}^{(2)} \equiv \sum_{\Delta\ne0}
  \f{\v w_\Delta\c\v w_\Delta}{\Delta},
  &&
  w_{\t{pairs}}^{(2)} \equiv \sum_{\Delta\ne0}
  \f{\v{\m w}_\Delta\c\v{\m w}_\Delta}{\Delta},
\end{align}
in terms of which
\begin{align}
  W_4 \equiv \f{w_{\t{columns}}^{(2)} - w_{\t{pairs}}^{(2)}}{N!_4},
  &&
  W_2 \equiv \f{w_{\t{columns}}^{(2)} - 2 w_{\t{pairs}}^{(2)}}
  {4N!_2} + \f12 \p{3N-4} W_4,
\end{align}
where
\begin{align}
  N!_k \equiv \prod_{j=0}^{k-1} \p{N-j} = \f{N!}{\p{N-k}!}
\end{align}
denotes a falling factorial.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Existence of a spectral gap with power-law interactions}
\label{sec:gap}

>>>>

We consider an array of $N$ multilevel spins with SU($n$)-symmetric
interactions that can be written in the form
\begin{align}
  H_0 = \sum_{p<q} h_{pq} \Pi_{pq},
  &&
  \Pi_{pq} \equiv \sum_{\mu,\nu} s_{\mu\nu}^{(p)} s_{\nu\mu}^{(q)},
  \label{eq:H_0}
\end{align}
where $p,q$ index individual spins; $\mu,\nu$ index states in an
orthonormal basis for the $n$-dimensional Hilbert space $\H_n$ of a
single spin; $h_{pq}$ are scalar coefficients; the operator
$s_{\mu\nu}^{(p)}\equiv\op{\mu}{\nu}_p$ flips the state of spin $p$ to
$\ket\mu$ from $\ket\nu$; and the operator $\Pi_{pq}$ permutes the
states of spins $p$ and $q$.  The permutation operator $\Pi_{pq}$ is a
multilevel generalization of the SU(2)-symmetric spin interaction
$\v s_p\c\v s_q$ for $\v s=\p{s_\x,s_\y,s_\z}$, with
$s_\alpha=\sigma_\alpha/2$ and $\sigma_\alpha$ a Pauli matrix for an
individual spin.

@@

If the coefficients $h_{pq}$ of the interaction Hamiltonian $H_0$ are
all negative, then the ground-state manifold $\M_0$ of the interaction
Hamiltonian $H_0$ consists of permutationally symmetric states that
are simultaneous $+1$ eigenstates of all permutation operators
$\Pi_{pq}$.  We will adopt the restriction that all $h_{pq}<0$
throughout these notes, but note that this restriction can be relaxed
to the assumption that the initial state of all spins is
permutationally symmetric (e.g.~a spin-polarized state).  We will also
assume throughout these notes that the permutationally symmetric
manifold $\M_0$ is gapped away from all other states by an interaction
energy difference that is large compared to any coupling between
$\M_0$ and its orthogonal complement.  Power-law couplings of the form
$h_{pq}=-h/\abs{p-q}^\alpha$ on a $D$-dimensional lattice, for
example, always yield a non-vanishing spectral gap when $\alpha\le D$
(see Appendix \ref{sec:spectral_gap}).

@@

To compute the effective Hamiltonian induced by the single-body
perturbation $\V_1$, we use the permutational symmetry of the
ground-state manifold to simplify
\begin{align}
  H_1^{(1)} = \sum_{X,p} v_{Xp} \P_0 X_p \P_0
  \EQPS \sum_{X,p} v_{Xp} X_1
  \EQPS \sum_X \bar v_X \col{X},
  \label{eq:H_1_1}
\end{align}
where $\EQPS$ denotes equality under a restriction to the
permutationally symmetric manifold $\M_0$, and
\begin{align}
  \bar v_X \equiv \f1N \sum_p v_{Xp},
  &&
  \col{X} \equiv \sum_p X_p,
\end{align}
are respectively the mean value in $\v v_X$ and the collective version
of $X$.  In order to compute the second order effective Hamiltonian
$H_1^{(2)}$, we first choose an arbitrary permutationally symmetric
state $\ket\psi\in\M_0$ and expand (see Appendix \ref{sec:H_V1_psi}):
\begin{align}
  H_0 \V_1 \ket\psi
  = E_0 \V_1 \ket\psi
  + \sum_X \sum_p \p{\sum_q h_{pq} v_{Xq} - h_p v_{Xp}} X_p
  \ket\psi,
  &&
  h_p \equiv \sum_q h_{pq},
  \label{eq:H_V1_psi}
\end{align}
where we define $h_{pq}$ for all $p,q$ (as opposed to only $p<q$) by
enforcing $h_{pq}=h_{qp}$ and $h_{pp}=0$.  We thus find that a
perturbation $\V_1^\Delta\ket\psi$ (indexed by $\Delta$) is an
eigenvector of the interaction Hamiltonian $H_0$ if its coefficient
vectors $\v v_X^\Delta$ all satisfy the eigenvalue equation
\begin{align}
  \p{\m h - \diag\v h}\c\v v = \Delta \v v,
  \label{eq:V1_eig}
\end{align}
where $\m h\equiv\sum_{p,q}h_{pq}\op{p}{q}$ is a matrix of all
couplings $h_{pq}$; the vector $\v h\equiv\sum_{p,q}h_{pq}\ket{p}$ is
the sum of all columns of $\m h$; and $\diag\v h$ is a matrix with
$\v h$ on the diagonal and zeroes everywhere else.  Solving the
general eigenvalue problem in \eqref{eq:V1_eig} requires diagonalizing
the $N\times N$ matrix $\m h-\diag\v h$.  When the interaction
Hamiltonian $H_0$ is translationally invariant, however, this
eigenvalue problem can be solved analytically; we provide the
corresponding solution at the end of Appendix \ref{sec:H_V1_psi}.

If we {\it construct} a perturbation $\V_1^\Delta$ with coefficients
$\v v_X^\Delta$ that satisfy \eqref{eq:V1_eig} with eigenvalue
$\Delta$, then
\begin{align}
  H_0 \V_1^\Delta \ket\psi = \p{E_0 + \Delta} \V_1^\Delta \ket\psi.
\end{align}
Interestingly, the energy of the state $\V_1^\Delta\ket\psi$ depends
only on the coefficients $\v v_X^\Delta$, and is entirely independent
of the state $\ket\psi$ or the choice of a trace-zero single-spin
operator $X$ used to build $\V_1^\Delta$.  Finding operators
$\V_1^\Delta$ that generate eigenvectors of the interaction
Hamiltonian $H_0$ when they act on permutationally symmetric states
$\ket\psi\in\M_0$ thus reduces to finding eigenvectors of the
interaction matrix $\m h-\diag\v h$.

Any coefficient vector $\v v_X$ can be expanded into its projections
$\v v_X^\Delta$ onto the eigenspace of vectors satisfying
\eqref{eq:V1_eig} with eigenvalue $\Delta$,
i.e.~$\v v_X = \sum_\Delta \v v_X^\Delta$, which also allows us to
expand
\begin{align}
  \V_1 = \sum_\Delta \V_1^\Delta,
  &&
  \V_1^\Delta \equiv \sum_X \v v_X^\Delta \c \v X,
\end{align}
where each operator $\V_1^\Delta$ generates a state with definite
interaction energy $\Delta$ above that of the permutationally
symmetric manifold $\M_0$.  We can therefore simplify
\begin{align}
  H_1^{(2)}
  = - \P_0 \V_1 \E \V_1 \P_0
  = - \sum_{\Delta\ne0} \f1{\Delta}
  \P_0 \V_1^\Delta \P_\Delta \V_1^\Delta \P_0
  = - \sum_{\Delta\ne0} \f1{\Delta} \P_0 \p{\V_1^\Delta}^2 \P_0,
\end{align}
where the product $\P_0 \p{\V_1^\Delta}^2 \P_0$ is worked out in
Appendix \ref{sec:PXYP}.  In total, we find that
\begin{align}
  H_1^{(2)}
  \EQPS \f1{N\p{N-1}} \sum_{X,Y} \sum_{\Delta\ne0}
  \f{\v v_X^\Delta\c\v v_Y^\Delta}{\Delta}
  \p{\col{X}\,\col{Y} - N \col{XY}},
  &&
  \col{Z} \equiv \sum_p Z_p.
  \label{eq:H_1_2}
\end{align}
where $\EQPS$ denotes equality under a restriction to the
permutationally symmetric manifold $\M_0$; and $\v v_X^\Delta$ is the
projection of $\v v_X$ onto the $\Delta$-eigenspace of the matrix
$\m h-\diag\v h$.

@@

The first-order effective Hamiltonian $H_2^{(1)}$ induced on the
permutationally symmetric manifold $\M_0$ by the two-body perturbation
$\V_2$ in \eqref{eq:perturbations} is simply the restriction of $\V_2$
onto $\M_0$:
\begin{align}
  H_2^{(1)} = \sum_O \sum_{p<q} w_{Opq} \P_0 O_{pq} \P_0
  \EQPS \sum_O \sum_{p<q} w_{Opq} O_{1,2}
  \EQPS \sum_O \bar w_O \col{O},
\end{align}
where
\begin{align}
  \bar w_O \equiv {N \choose 2}^{-1} \sum_{p<q} w_{Opq},
  &&
  \col{O} \equiv \sum_{p<q} O_{pq},
\end{align}
are an average of $w_{Opq}$ and a collective version of the two-spin
operator $O_{pq}$.  If the two-spin operator
$O=X\otimes Y+Y\otimes X$, then
\begin{align}
  H_2^{(1)} \EQPS \f12 \sum_{\p{X,Y}} \bar w_{\p{X,Y}}
  \p{\col{X}\,\col{Y} + \col{Y}\,\col{X} - \col{XY} - \col{YX}}.
\end{align}
In order to compute the second-order effective Hamiltonian
$H_2^{(2)}$, as before we pick an arbitrary permutationally symmetric
state $\ket\psi\in\M_0$ and expand (see Appendix
\ref{sec:eigenstates})
\begin{align}
  H_0 \V_2 \ket\psi
  = E_0 \V_2 \ket\psi
  + \sum_O \sum_{p<q}
  \sp{\sum_k \p{h_{pk} w_{Okq} + h_{qk} w_{Opk}}
  - \bar h_{pq} w_{Opq}}
  O_{pq} \ket\psi,
  \label{eq:H_V2_psi}
\end{align}
where to keep notation compact, we define
\begin{align}
  \bar h_{pq} \equiv h_p + h_q - 2 h_{pq},
  &&
  h_p \equiv \sum_q h_{pq},
\end{align}
and we define $w_{Opq}$ for all $p,q$ (as opposed to only $p<q$) by
enforcing $w_{Opq}=w_{Oqp}$ and $w_{Opp}=0$.  We thus find that the
vector $\V_2^\Delta\ket\psi$ (indexed by $\Delta$) is an eigenvector
of the interaction Hamiltonian $H_0$ if its coefficients
$\v{\m w}_O^\Delta$ all satisfy the eigenvalue equation
\begin{align}
  \check{\m h} \c \v{\m w} = \Delta \v{\m w},
  \label{eq:V2_eig}
\end{align}
where
\begin{align}
  \check{\m h}
  \equiv \sum_{p<q} \ket{\set{p,q}} \sp{\sum_{k\notin\set{p,q}}
    \p{h_{pk} \bra{\set{k,q}} + h_{qk} \bra{\set{p,k}}}
    - \bar h_{pq} \bra{\set{p,q}}}.
  \label{eq:V2_eig_mat}
\end{align}
We explicitly exclude $k\in\set{p,q}$ in \eqref{eq:V2_eig_mat} because
these terms vanish in \eqref{eq:H_V2_psi} from the fact that
$h_{kk}=w_{Okk}=0$.  Solving the general eigenvalue problem in
\eqref{eq:V2_eig} requires diagonalizing the
${N \choose 2}\times{N \choose 2}\sim N^2\times N^2$ matrix
$\check{\m h}$.  When both he interaction Hamiltonian $H_0$
translationally invariant, however, this eigenvalue problem can be
reduced to that of diagonalizing an $\sim N\times N$ matrix; we
perform this reduction in Appendix \ref{sec:eigenstates_TI}.

If we {\it construct} a perturbation $\V_2^\Delta$ with coefficients
$\v{\m w}_O^\Delta$ that satisfy \eqref{eq:V2_eig} with eigenvalue
$\Delta$, then
\begin{align}
  H_0 \V_2^\Delta \ket\psi = \p{E_0 + \Delta} \V_2^\Delta \ket\psi,
\end{align}
which implies that $\V_2^\Delta\ket\psi$ is a state of definite energy
that depends only on the coefficients $\m w_O^\Delta$.  Similarly to
the case of single-body perturbations, we can decompose any vector
$\v{\m w}_O$ into its projections $\v{\m w}_O^\Delta$ onto the
eigenspace of vectors satisfying \eqref{eq:V2_eig} with eigenvalue
$\Delta$, i.e.~$\v{\m w}_O=\sum_\Delta\v{\m w}_O^\Delta$, which allows
us to expand
\begin{align}
  \V_2 = \sum_\Delta \V_2^\Delta,
  &&
  \V_2^\Delta \equiv \sum_O \v{\m w}_O^\Delta \c \v{\m O},
\end{align}
and in turn write
\begin{align}
  H_2^{(2)} = - \P_0 \V_2 \E \V_2 \P_0
  = -\sum_{\Delta\ne0} \f1\Delta \P_0 \p{\V_2^\Delta}^2 \P_0,
\end{align}
where the product $\P_0 \p{\V_2^\Delta}^2 \P_0$ is simplified in
Appendix \ref{sec:POQP}.  In the special case that
\begin{align}
  \V_2 = \sum_{p<q} w_{pq} s_\z^{(p)} s_\z^{(q)},
\end{align}
where $s_\z\equiv\p{\op\up-\op\dn}/2$ is an SU(2) spin-$z$ operator,
the second-order effective Hamiltonian $H_2^{(2)}$ takes the form
\begin{align}
  H_2^{(2)} \simeq_\PS W_4 S_\z^4 - W_2 S_\z^2,
\end{align}
where $\simeq_\PS$ denotes equality up to scalar terms and a
restriction to the permutationally symmetric manifold;
$S_\z\equiv\sum_p s_\z^{(p)}$ is a collective SU(2) spin operator; and
$W_k$ are scalar coefficients.  In order to expand the coefficients
$W_k$, we define
\begin{align}
  \v{\m w}=\sum_{p<q} w_{pq} \ket{\set{p,q}},
  &&
  \v w \equiv \sum_{p,q} w_{pq} \ket{p},
\end{align}
where $w_{pq}$ is defined for all $p,q$ (as opposed to only $p<q$) by
enforcing $w_{pq}=w_{qp}$ and $w_{pp}=0$.  The vector $\v w$ is the
sum of all columns of the matrix
$\m w\equiv\sum_{p,q}w_{pq}\op{p}{q}$.  We then denote the projection
of $\v{\m w}$ onto the $\Delta$-eigenspace of the matrix
$\check{\m h}$ defined in \eqref{eq:V2_eig_mat} by $\v{\m w}_\Delta$,
and define
\begin{align}
  w_{\t{columns}}^{(2)} \equiv \sum_{\Delta\ne0}
  \f{\v w_\Delta\c\v w_\Delta}{\Delta},
  &&
  w_{\t{pairs}}^{(2)} \equiv \sum_{\Delta\ne0}
  \f{\v{\m w}_\Delta\c\v{\m w}_\Delta}{\Delta},
\end{align}
in terms of which
\begin{align}
  W_4 \equiv \f{w_{\t{columns}}^{(2)} - w_{\t{pairs}}^{(2)}}{N!_4},
  &&
  W_2 \equiv \f{w_{\t{columns}}^{(2)} - 2 w_{\t{pairs}}^{(2)}}
  {4N!_2} + \f12 \p{3N-4} W_4,
\end{align}
where
\begin{align}
  N!_k \equiv \prod_{j=0}^{k-1} \p{N-j} = \f{N!}{\p{N-k}!}
\end{align}
denotes a falling factorial.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Existence of a spectral gap with power-law interactions}
\label{sec:spectral_gap}

Here we show that power-law SU($n$)-symmetric interactions of the form
in \eqref{eq:H_0} with $h_{pq}=-h/\abs{p-q}^\alpha$ yield a
non-vanishing many-body interaction energy gap when $\alpha\le D$,
where $D$ is the dimension of the lattice.  On a periodic lattice of
$N=L^D$ spins, the spectral gap of the interaction Hamiltonian is (see
Appendix \ref{sec:H_V1_psi})
\begin{align}
  \Delta
  = \sum_{d\in\ZZ_L^D} h_{0,d} \sp{\cos\p{d \c k_{\t{SE}}}-1}
  = h \sum_{\substack{d\in\ZZ_L^D\\\abs{d}\ge1}}
  \f{1-\cos\p{d \c k_{\t{SE}}}}{\abs{d}^\alpha},
\end{align}
where the domain of $d$ is defined in terms of integers modulo $L$,
i.e.~$\ZZ_L\simeq\set{0,1,\cdots,L-1}$ with the relation $\simeq$
denoting an association that ignores the cyclic structure of $\ZZ_L$;
$k_{\t{SE}}\in\ZZ_L\times2\pi/L$ is a wavenumber for a singly-excited
spin-wave state; and the norm $\abs{d}$ for a vector $d$ on a periodic
lattice is implicitly understood to mean the smallest Euclidean
distance of $d$ from a fixed origin.  In order to yield the smallest
excitation energy $\Delta$, the wavenumber $k_{\t{SE}}$ should
maximize the contribution of the cosine term above, which is achieved
by a wavenumber that minimizes the oscillations of this term when
integrated over the entire lattice.  A suitable candidate for a
minimal wavenumber is $k_{\t{SE}}=\p{2\pi/L,0,0,\cdots}$, which
corresponds to an excitation energy
\begin{align}
  \Delta = h \sum_{\substack{d\in\ZZ_L^D\\\abs{d}\ge1}}
  \f{1-\cos\p{d_12\pi/L}}{\abs{d}^\alpha}.
\end{align}
Defining $\epsilon\equiv2/L$ and a rescaled domain symmetrized about
$0$, $\SS_\epsilon\simeq\ZZ_L/\epsilon$ with
$\SS_\epsilon\subset\sp{-1,1}$, we substitute $x\simeq\epsilon d$ to
get
\begin{align}
  \Delta
  = h \sum_{\substack{x\in\SS_\epsilon^D\\\abs{x}\ge\epsilon}}
  \f{1-\cos\p{\pi x_1}}{\abs{x/\epsilon}^\alpha}
  = h \epsilon^{\alpha-D} \sum_{\substack{x\in\SS_L^D\\\abs{x}\ge\epsilon}}
  \epsilon^D \f{1-\cos\p{\pi x_1}}{\abs{x}^\alpha}.
\end{align}
As $\epsilon\to0$ the discrete sum over $x$ becomes an integral that
avoids an infinitesimal region about the origin, i.e.
\begin{align}
  \Delta = h \epsilon^{\alpha-D} \I_{D\epsilon},
  &&
  \I_{D\epsilon}
  \equiv \int_{\TT_1^D\setminus\TT_\epsilon^D} \d^Dx\,
  \f{1-\cos\p{\pi x_1}}{\abs{x}^\alpha},
\end{align}
where the integral $\I_{D\epsilon}$ is defined using the interval
$\TT_a\equiv\p{-a,a}$.  The integrand of $\I_{D\epsilon}$ is strictly
positive and well-behaved on the entirety of its domain except for the
origin, where depending on the value of $\alpha$ the integrand may
vanish or diverge as $\abs{x}\to0$.  Together, these facts mean that
\begin{align}
  \I_{D\epsilon} \stackrel{\epsilon\to0}{\sim} \epsilon^{-\gamma},
  &&
  \Delta \stackrel{\epsilon\to0}{\sim} h \epsilon^{\alpha-D-\gamma},
\end{align}
for some $\gamma\ge0$, which implies that gap $\Delta$ is
non-vanishing when $\alpha\le D\le D+\gamma$.
